\Large \bf Cyclone: High Availability for Persistent Key Value Stores

### 2017-08-07 paper/paper.tex
Minor updates

\begin{abstract}
\vspace{0.1in}
Persistent key value stores are rapidly becoming an important component of many
distributed data serving solutions with innovations targeted at taking advantage
of growing flash speeds. Unfortunately their performance is hampered by the need
to maintain and replicate a write ahead log to guarantee availability in the
face of machine and storage failures. Cyclone is a replicated log plug-in for
key value stores that systematically addresses various sources of this
bottleneck. It uses a small amount of non-volatile memory directly addressable
by the CPU - such as in the form of NVDIMMs - to remove block oriented IO
devices such as SSDs from the critical path for appending to the log. It then
addresses network overheads using an implementation of the RAFT consensus
protocol that is designed around a userspace network stack to relieve the CPU of
the burden of data copies. Finally, it provides a way to efficiently map the
commutativity in key-value store APIs to the parallelism available in commodity
NICs via multiple transmit receive queues to scale replication performance on
multicore machines. Cyclone is able to replicate millions of small updates a
second using only commodity 10 gigabit ethernet adapters. As a practical
application, we use it to improve the performance and availability of RocksDB, a
popular persistent key value store by an order of magnitude when compared to its
own write ahead log without replication.
\end{abstract}

\section{Introduction}
Persistent key value stores are an increasingly important component of
datacenter scale storage services. Key value stores such as
Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
large amounts of effort on both the engineering and research fronts. These key
value stores include sophisticated in-memory data structures built around Log
Structured Merge trees~\cite{lsmtree} and are heavily tuned to extract maximum
performance from flash-based solid state drives (SSDs).

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems - eclipsing
much of the work on improving the performance of the LSM component. To
illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}, depending on upper layers to provide
availability by co-ordinating replicas. On the other side of the spectrum, key
value store research prototypes such as FloDB~\cite{flodb} turn off the write
ahead log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

Cyclone avoids this by making use of directly attached
persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
an added ultracapacitor to dump state to a small amount of attached flash) or
newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
are significantly more expensive than standard DIMMs. This means that they will
likely be available in small quantities. We compensate for this by periodically
draining the NVDIMM log into a log placed on a standard SSD on the system.

The second key contribution of this work is in showing that the throughput
bottlenecks in the network stack can be mitigated entirely in software. Existing
work on log replication using state machine replication protocols such as
Paxos~\cite{paxos} and RAFT~\cite{raft} has not - thus far - addressed high
performance replication in the local area network of a datacenter using
commodity networks.  Research focusing on addressing wide area network latencies
by reducing the number of network hops with new protocols such as Fast
Paxos~\cite{fast-paxos} or network switch modifications such as No
Paxos~\cite{nopaxos} do not address throughput bottlenecks due to network stack
related inefficiencies in the participating nodes of the quorum - a more urgent
problem with local area replication. On the other hand, work done to address
network related inefficiencies start with the assumption that the bottleneck to
good throughput on the network is the CPU and therefore one must either resort
to offload mechanisms such as RDMA~\cite{dare, farm, faast} or offload the
protocol entirely to an FPGA~\cite{consensus_box}. We demonstrate in this paper
that this is not the case. Rather, once persistence is provided by directly
attached non-volatile memory eliminating block IO from the persistence step in
replication protocols, the problem becomes equivalent to multicasting the same
log entry from the leader to all followers. We design Cyclone's network stack
around well known principles in the software packet switching community. In
particular, a careful implementation of a consensus protocol that relieves the
CPU of the responsibility of data movement permits a high performance
implementation entirely in software that addresses the order of magnitude
network related performance gap in Figure~\ref{fig:problem} using only commodity
10 Gigabit ethernet.

@@

A key value store typically serves multiple client requests concurrently and is
linearizable~\cite{linearizability}. This means any history of requests and
responses can be rearranged such that responses immediately follow corresponding
requests and if a response precedes a request in the original history it also
does so in the rearranged history. Key value stores such as RocksDB also
support atomic writes ({\tt PUT} operations) to multiple keys, called batched
writes and the ability to take snapshots.

Cyclone integrates with key value stores as both a client and server side
library. Cyclone replicates the key value store across a set of replica
servers. The client side library sends \emph{all} requests to a distinguished
leader replica.  On the server side Cyclone accepts requests from clients and calls
into the key value store. It then returns a response to the client. Before
executing any PUT request, Cyclone appends it to a durable log and replicates
the request to the logs of follower replicas. The request is considered
replicated once acknowledged by a majority of replicas. Follower replicas apply
updates from the log in order.

Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On
failure of the leader replica, a new leader is automatically elected - RAFT
ensures that the new leader has the most up to date log. The Cyclone client
library automatically locates the new leader on a failover - clients therefore
see only a brief interruption in service. Although read requests (GET
operations) are not logged, they are serviced in FIFO order with the write
requests (PUT/DELETE operations) currently in Cyclone's log at the moment the
read request is received. This ensures linearizability in the face of failovers.

\section{Storage}
\label{sec:storage}
Storage devices such as flash-based SSDs export a block IO interface. Appending
a small update to the write-ahead log requires a synchronous access to the
SSD. Even with write coalescing using a write cache on the SSD, the round trip
time to a current generation NVMe SSD such as the one we use in this
paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
under 50K ops/sec. This leads to a high baseline latency at even low load, as
shown in Figure~\ref{fig:problem}. This problem leads many system designers to
underprovision key-value store shards ensuring that it operates at moderate to
high load, using group-commit to batch additions to the log. They therefore pay
a price in latency to ensure the bandwidth to storage is fully utilized.

Cyclone adopts a different strategy by making use of novel memory technologies
to obviate the need to make this tradeoff in the first place. We assume a small
amount of non-volatile memory directly addressable by the CPU rather than being
placed behind a block oriented IO interface. Log appends are done to the ``NVM
log'' placed in this non-volatile memory. While we designed Cyclone to leverage
new non-volatile memory technologies that might be available in the
future~\cite{3dxpoint}, current non-volatile memory options such are NVDIMMs are
more expensive that DRAM. Therefore, for Cyclone to be economical we cannot
assume that a sufficient amount of NVM is available to hold the entire log. The
log entries can only be garbage collected when the in-memory component of the
key value store is checkpointed to secondary storage. The growing size of
available DRAM means that also provisioning sufficient NVDIMMs to hold the
corresponding log of updates is not feasible given that NVDIMMs are
significantly more expensive than standard volatile DIMMs.  To illustrate this,
at the time of writing this paper, an 8 GB DDR4 NVDIMM was retailing at \$612,
while a plain 8 GB DD4 DIMM was retailing at \$178.  We therefore resort to
using a second larger log level placed on a flash SSD. Entries are drained from
the NVM log to this flash log in conveniently large units that are a multiple of
4KB - the optimum IO unit for flash SSDs. This ensures we get the best possible
throughput from a flash device without paying the price of synchronous IO for
small key-value pairs. At the same time, the amount of non-volatile memory
required is small enough to not add undue cost of the server.

>>>>

\begin{abstract}
\vspace{0.1in}
Persistent key value stores are an important component of many
distributed data serving solutions with innovations targeted at taking advantage
of growing flash speeds. Unfortunately their performance is hampered by the need
to maintain and replicate a write ahead log to guarantee availability in the
face of machine and storage failures. Cyclone is a replicated log plug-in for
key value stores that systematically addresses various sources of this
bottleneck. It uses a small amount of non-volatile memory directly addressable
by the CPU - such as in the form of NVDIMMs - to remove block oriented IO
devices such as SSDs from the critical path for appending to the log. It then
addresses network overheads using an implementation of the RAFT consensus
protocol that is designed around a userspace network stack to relieve the CPU of
the burden of data copies. Finally, it provides a way to efficiently map the
commutativity in key-value store APIs to the parallelism available in commodity
NICs via multiple transmit receive queues to scale replication performance on
multicore machines. Cyclone is able to replicate millions of small updates a
second using only commodity 10 gigabit ethernet adapters. As a practical
application, we use it to improve the performance and availability of RocksDB, a
popular persistent key value store by an order of magnitude when compared to its
own write ahead log without replication.
\end{abstract}

\section{Introduction}
Persistent key value stores are an important component of
datacenter scale storage services. Key value stores such as
Rocksdb~\cite{rocksdb}, LevelDB~\cite{leveldb} and FloDB~\cite{flodb} represent
significant efforts on both engineering and research fronts. These key
value stores include sophisticated in-memory data structures built around Log
Structured Merge trees~\cite{lsmtree} and are heavily tuned to extract maximum
performance from flash-based solid state drives (SSDs).

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems, eclipsing
much of the work on improving the performance of the LSM component. To
illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}, depending on upper layers to provide
availability by co-ordinating replicas. On the other side of the spectrum, key
value store research prototypes such as FloDB~\cite{flodb} turn off the write
ahead log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

Cyclone avoids this overhead by making use of directly attached
persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
an added ultracapacitor to dump state to a small amount of attached flash) or
newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
are significantly more expensive than standard DIMMs, which means that they will
likely be available only in smaller quantities compared to flash. We compensate 
for NVDIMM's size by periodically draining the NVDIMM log into a log placed on a
standard SSD on the system.

The second key contribution of this work is in showing that the throughput
bottlenecks in the network stack can be mitigated entirely in software. Existing
work on log replication using state machine replication protocols such as
Paxos~\cite{paxos} and RAFT~\cite{raft} has not - thus far - addressed high
performance replication in the local area network of a datacenter using
commodity networks.  Research focusing on addressing wide area network latencies
by reducing the number of network hops with new protocols such as Fast
Paxos~\cite{fast-paxos} or network switch modifications such as No
Paxos~\cite{nopaxos} do not address throughput bottlenecks due to network stack
related inefficiencies in the participating nodes of the quorum -- a more urgent
problem with local area replication. On the other hand, work done to address
network related inefficiencies start with the assumption that the bottleneck to
good throughput on the network is the CPU and therefore one must either resort
to offload mechanisms such as RDMA~\cite{dare, farm, faast} or offload the
protocol entirely to an FPGA~\cite{consensus_box}. We demonstrate in this paper
that this is not the case. Rather, once persistence is provided by directly
attached non-volatile memory eliminating block IO from the persistence step in
replication protocols, the problem becomes equivalent to multicasting the same
log entry from the leader to all followers. We design Cyclone's network stack
around well known principles in the software packet switching community. In
particular, a careful implementation of a consensus protocol that relieves the
CPU of the responsibility of data movement permits a high performance
implementation entirely in software that addresses the order of magnitude
network related performance gap in Figure~\ref{fig:problem} using only commodity
10 Gigabit ethernet.

@@

A key value store typically serves multiple client requests concurrently and is
linearizable~\cite{linearizability}. This means any history of requests and
responses can be rearranged such that responses immediately follow corresponding
requests and if a response precedes a request in the original history it also
does so in the rearranged history. Key value stores such as RocksDB also
support atomic writes ({\tt PUT} operations) to multiple keys, called batched
writes, and the ability to take snapshots.

Cyclone integrates with key value stores as both a client and server side
library. Cyclone replicates the key value store across a set of replica
servers. The client side library sends \emph{all} requests to a distinguished
leader replica.  On the server side Cyclone accepts requests from clients, calls
into the key value store, and returns a response to the client. Before
executing any PUT request, Cyclone appends it to a durable log and replicates
the request to the logs of follower replicas. The request is considered
replicated once acknowledged by a majority of replicas. Follower replicas apply
updates from the log in order.

Cyclone uses RAFT protocol~\cite{raft} to keep the logs in sync across the
replicas. On
failure of the leader replica, a new leader is automatically elected -- RAFT
ensures that the new leader has the most up to date log. The Cyclone client
library automatically locates the new leader on a failover, resulting in only a
brief interruption in service for clients. Although read requests (GET
operations) are not logged, they are serviced in FIFO order with the write
requests (PUT/DELETE operations) in Cyclone's log at the moment the
read request is received. This ensures linearizability in the face of failovers.

\section{Storage}
\label{sec:storage}
Storage devices such as flash-based SSDs export a block IO interface. Appending
a small update to the write-ahead log requires a synchronous access to the
SSD. Even with write coalescing using a write cache on the SSD, the round trip
time to a current generation NVMe SSD such as the one we use in this
paper~\cite{ssd_spec}, is of the order of 20 us, limiting the throughput to
under 50K ops/sec and resulting in a high baseline latency at even low load (as
shown in Figure~\ref{fig:problem}). This problem leads many system designers to
underprovision key-value store shards ensuring that it operates at moderate to
high load, using group-commit to batch additions to the log. They therefore pay
a price in latency to ensure the bandwidth to storage is fully utilized.

Cyclone adopts a different strategy by making use of novel memory technologies
to obviate the need to make this tradeoff in the first place. We assume a small
amount of non-volatile memory directly addressable by the CPU rather than being
placed behind a block oriented IO interface. Log appends are done to the ``NVM
log'' placed in this non-volatile memory.
%% TODO: Messaging has to be improved here. It appears as if the problems
%% with NVDIMM (as far as Cyclone is concerned) will be solved once 3DXPoint
%% appears -- lower cost and higher capacity than DRAM.
While we designed Cyclone to leverage
new non-volatile memory technologies that might be available in the
future~\cite{3dxpoint}, current non-volatile memory options such are NVDIMMs are
more expensive that DRAM. Therefore, for Cyclone to be economical we cannot
assume that a sufficient amount of NVM is available to hold the entire log. The
log entries can only be garbage collected when the in-memory component of the
key value store is checkpointed to secondary storage. The growing size of
available DRAM means that also provisioning sufficient NVDIMMs to hold the
corresponding log of updates is not feasible given that NVDIMMs are
significantly more expensive than standard volatile DIMMs.  To illustrate this,
at the time of writing this paper, an 8 GB DDR4 NVDIMM was retailing at \$612,
while a plain 8 GB DD4 DIMM was retailing at \$178.  We therefore resort to
using a second larger log level placed on a flash SSD. Entries are drained from
the NVM log to this flash log in conveniently large units that are a multiple of
4KB - the optimum IO unit for flash SSDs. This ensures we get the best possible
throughput from a flash device without paying the price of synchronous IO for
small key-value pairs. At the same time, the amount of non-volatile memory
required is small enough to not add undue cost to the server.
